Unreliable heterogeneity: how measurement error obscures heterogeneity in meta-analyses in psychology

### 2023-06-18 manuscript/manuscript.qmd
updated abstract

---
title: "Unreliable heterogeneity: how measurement error obscures heterogeneity in meta-analyses in psychology"
date: today
date-format: long
author:
  - name: Anton Olsson-Collentine
    url: https://example.com
    orcid: 0000-0000-0000-0000
    affiliations:
      - id: tiu
        name: "Department of Methodology and Statistics, Tilburg School of Social and Behavioral Sciences, Tilburg University, the Netherlands."
    email: anton@olssoncollentine.com
  - name: Marjan Bakker
    affiliations:
      - ref: tiu
  - name: Jelte Wicherts
    affiliations:
      - ref: tiu
format:
  bj-pdf:
    classoption:
      - twocolumn
colorlinks: true
link-citations: true
keep-tex: false
abstract: |
  One of the main purposes of meta-analysis is to estimate heterogeneity of effect sizes. The quality of meta-analytic estimates depends in large part on the quality of evidence in primary studies, yet there exist broad concerns that proper measurement tends to be neglected in much of psychological primary research. Such inattention to measurement can also bias heterogeneity estimates. We performed a Monte-Carlo simulation study to examine the expected bias in heterogeneity estimates due to unreliability in primary studies. Although unreliability in primary studies can both inflate and suppress heterogeneity, our results indicate that under most circumstances unreliability in primary studies leads to a severe underestimate of heterogeneity in meta-analyses. For a typical effect size heterogeneity can be expected to be underestimated by 15-60%, or more if average reliability in primary studies is below 0.6. We observed a positive bias in heterogeneity estimates due to unreliability only under specific and arguably uncommon circumstances of (1) actual zero heterogeneity, particularly when mean effect sizes are large, or (2) combinations of very small true heterogeneity, large variance in primary study reliabilities, large mean effect sizes, and a limited number of primary studies. The severity of the underestimation of heterogeneity could lead to meta-analyses in psychology missing relevant moderators and overgeneralizing outcomes. Unreliability appears to bring truly different effects closer together, thereby obscuring true differences between studies that could be relevant for theory, practice, and future research efforts. Research on concrete guidance to applied meta-analysts is urgently needed, as sophisticated methods for correcting measurement unreliability such as meta-analytic structural equation modeling (MASEM) are only applicable in exceptional cases and corrections based on classical test theory come with caveats and strong assumptions.
license: CC BY
keywords:
  - Research
  - Cool Stuff
citation:
  type: article
  citation-key: citationKeyNumber
  status: draft
cite-method: citeproc
csl: apa.csl
bibliography: [unreliable-heterogeneity.bib]
---

>>>>

---
title: "Unreliable heterogeneity: how measurement error obscures heterogeneity in meta-analyses in psychology"
date: today
date-format: long
author:
  - name: Anton Olsson-Collentine
    url: https://example.com
    orcid: 0000-0000-0000-0000
    affiliations:
      - id: tiu
        name: "Department of Methodology and Statistics, Tilburg School of Social and Behavioral Sciences, Tilburg University, the Netherlands."
    email: anton@olssoncollentine.com
  - name: Marjan Bakker
    affiliations:
      - ref: tiu
  - name: Jelte Wicherts
    affiliations:
      - ref: tiu
format:
  bj-pdf:
    classoption:
      - twocolumn
colorlinks: true
link-citations: true
keep-tex: false
abstract: |
  Measurement error (imperfect reliability) is present in any empirical effect size estimate and systematically attenuates observed effect sizes compared to true underlying effect sizes. Yet there exist broad concerns that proper measurement tends to be neglected in much of psychological research. We examined how measurement error in primary studies affects meta-analytic heterogeneity estimates using Monte-Carlo simulations. Our results indicate that although measurement error in primary studies can both inflate and suppress heterogeneity, under most circumstances measurement error in primary studies leads to a severe underestimate of heterogeneity in meta-analysis. Our simulations showed expected heterogeneity to be underestimated by about 15% - 60% when considering a typical effect size around r = 0.2 and true heterogeneity levels that are common in the meta-analytic literature (τ>0.1, in Pearson’s r). The underestimate primarily depends on average reliability in primary studies (higher reliability leads to a smaller underestimate), but also worsens with smaller primary study sample sizes. We observed a positive bias in heterogeneity estimates due to measurement error only under specific and arguably uncommon circumstances of (1) actual zero heterogeneity, particularly when mean effect sizes are large, or (2) combinations of very small true heterogeneity, large variance in primary study reliabilities, large mean effect sizes, and a limited number of primary studies. Severe underestimates of heterogeneity due to measurement error may affect many meta-analyses in psychology and obscure true differences between studies that could be relevant for theory, practice, and future research efforts. Research on concrete guidance to applied meta-analysts is needed, as sophisticated methods for correcting measurement unreliability such as meta-analytic structural equation modeling (MASEM) are only applicable in exceptional cases and corrections based on classical test theory come with caveats and strong assumptions.
license: CC BY
keywords:
  - Research
  - Cool Stuff
citation:
  type: article
  citation-key: citationKeyNumber
  status: draft
cite-method: citeproc
csl: apa.csl
bibliography: [unreliable-heterogeneity.bib]
---
